Disclosing funding sources for open access publication fees: the Open APC

### 2016-05-31 article.Rmd
streamline intrduction first try

Publication fees, often paid by funders or universities, are a widely discussed open access business model. Yet, how and to what extent these activities are effective in terms of the number of supported research articles and associated costs remains under debate. This paper describes the Open APC initiative[^2], in which German universities and research organization share spendings on open access publication fees, and how it is currently implemented. More specifically, it addresses three problem areas when studying the economical effects of open access publishing: fragmentation of open access funding, variable pricing schemes and the comparison across research institutions. Such an approach extends methods and improves data collection activities for researchers and practitioners, as well as contribute to a better understanding of factors affecting the analysis of publication fees in open access publishing.

The rise of open access journals  matches the increasing relevance of publication fees in academic publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. To cover these fees, authors tend to make use of funding that grant agencies or academic institutions provide [@Suber_2012]. However, collecting information about what was funded is in most cases difficult. One reason why payments made for open access journal publications are often hard to track is that, on the one hand, they are fragmented across the budgets of funding agencies, research institutions, and libraries, and, on the other hand, taken from personal budgets. Furthermore, open access funding mostly exists in higher income countries, mainly to support research articles in the bio- and physical sciences [@Solomon_2011]. Personal budgets stand in contrast with those support structures and are likely used to cover low price publication fees [@Solomon_2011; @Bj_rk_2015]. Along with the fragmentation of payments, funding for open access publications lacks transparency because the parties involved - authors, universities, funders, publishers - neither release information on who pays for what nor the costs of publishing [@Bj_rk_learned_2014], a situation similar to the lack of transparency regarding journal subscriptions [@Lawson_Meghreblian_2015]. It also remains unclear which factors contribute to price formation.[^4b] While fixed prices for individual articles are common, agreements between publishers and institutions often provide discounts and publishers sometimes waive publication fees for authors from low-income countries [@BJ_RK_2012; @Lawson_waiver_2015]. Other factors leading to a complex landscape of variable pricing schemes [@Pinfield_2015] include submission or page charges [@BJ_RK_2012]. Hybrid journals substantially add to this complexity, because comprehensive offset systems to avoid paying for the same article twice, through subscription and publication fee, are rare, which, in turn, leads to the phenomena of "double dipping" in scholarly publishing [@Pinfield_2015].

This complex situation of fee-based open access publishing creates difficulties for researchers and practitioners alike. Because of fragmented payments, the extent of funding remains unclear. To increase transparency, some research funders have begun collecting and disclosing expenditures for open access journal articles as open data. As per definition, open data is data that "can be freely used, modified, and shared by anyone for any purpose" [@OpenDataHandbook]. Therefore, opening up information about the funding of open access journal articles promises to enhance the discussion about current and future business models in academic publishing. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc] and the Austrian Science Fund FWF [@fwf_apc], who both released data on publication fees they had funded. The British not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report expenditures along with bibliographic information, including title, journal and publisher, persistent identifier to the publisher's version, and a link to a deposit in a subject repository. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates. In the case of the Wellcome Trust, crowd-sourcing data cleaning activities through a Google spreadsheet in combination with checks against bibliographic sources massively improved the spending data (see comments in @wellcome_apc).

The open access landscape in Germany, which is the focus of this paper, shares the general problems of in-transparency regarding funding schemes and costs as discussed above. The Deutsche Forschungsgemeinschaft (DFG), the largest research funder in Germany, has been encouraging open access publishing since years. It launched its "Open-Access Publishing" program in 2009 that has strongly influenced the support of open access publication fees through funds managed by university libraries.[^3] With this program, the DFG aims to help universities to establish support structures for publishing in open access journals where authors are requested to pay a publication fee. To reduce administrative burdens, grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. Examples include central invoicing schemes and related agreements between university libraries and publishers. Grantees are also required to report the institutional publication output and their fees paid for open access journal articles to the DFG on a regular basis, and to present the university-wide strategy to sustain the funds when DFG’s initial support runs out lately in 2019. The DFG enforces a set of criteria grantees have to comply with, leading to similar implementations for supporting open access publishing across German universities: these criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 € (excluding VAT) [@Fournier_2013]. Research institutes organized in the Fraunhofer-Gesellschaft, Helmholtz-Gemeinschaft, Leibniz-Gemeinschaft, and Max-Planck-Gesellschaft are not eligible for this funding program, contributing to the diversity of schemes in Germany. In response, some organizations have adopted similar processes to support authors. The Max-Planck-Gesellschaft operates their long-lasting open access activities, including handling spending and publisher agreements centrally, through the Max Planck Digital Library [@Schimmer_2013; @Sikora_2015], while the Leibniz-Gemeinschaft set up a dedicated open access fund in 2016.

The growing share of articles published in fee-based open access journals in recent years has led to calls for an unified approach towards funding of publication fees. The Allianz der Wissenschaftsorganisationen[^10c], representing all major research organizations in Germany, thus marks transparency as a major means to sustain an "adequate open access publication system" [@allianz]. However, there are various ways to achieve this goal. The existing approaches in Austria and the United Kingdom have one institution in charge to collect and analyze the data. The history of the Open APC initiative is rather bottom-up: In May 2014, Bielefeld University Library began to share its expenditures for publication fees. The library put its approach to the working group "Electronic Publishing" of the Deutsche Initiative für Netzwerkinformation (DINI)[^11] as a basis for discussion, and invited others to participate. Reflecting the increasing demand for publicly available data, contributions from Universität Regensburg and Universität Hannover followed soon after. As of writing, 29 universities and research institutes voluntary reported their data to the Open APC initiative to be included into a unified data-set of all expenditures.

The aim of this study was to examine how much German universities and research organisations spent on open access publication fees until 2015. Drawing on self-reported data from the Open APC initiative, the analysis focused on the amount of institutional payments for publication fees, and compared these findings with those from related Austrian and UK initiatives. We also asked how thoroughly self-reported articles were indexed in CrossRef, a DOI minting agency for scholarly literature, and analysed how institutional spending per articles was distributed over publishers and journal titles.

>>>>

<!-- Publication fees, often paid by funders or universities, are a widely discussed business model for open access journals. Yet, how and to what extent these activities are effective in terms of the number of supported research articles and associated costs remains under debate. This study contributes to a growing empirical base about institutional spending on open access publication fees. More specifically, it uses self-reported cost data from German universities and research organisations to examine how much they spent, and how the so funded open access journal articles were distributed over publishers and journal titles using bibliographic information from CrossRef. Such an approach extends methods and improves data collection activities for researchers and practitioners alike, as well as contribute to a better understanding of factors affecting the analysis of publication fees in open access publishing. -->

The rise of open access journals matches the increasing relevance of publication fees in academic publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. To cover these fees, authors tend to make use of funding that grant agencies or academic institutions provide [@Suber_2012]. However, collecting information about what was funded is in most cases difficult. One reason why payments made for open access journal publications are often hard to track is that they are fragmented across the budgets of funding agencies, research institutions, and libraries, or, if no institutional support exists, are taken from personal budgets. Institutional open access funding mostly exists in higher income countries, mainly to support research articles in the bio- and physical sciences [@Solomon_2011]. Personal budgets stand in contrast with those support structures and are likely used to cover low price publication fees [@Solomon_2011; @Bj_rk_2015].

Along with the fragmentation of payments, another problem is that funding for open access publications lacks transparency because the parties involved - authors, universities, funders, publishers - neither release information on who pays for what nor the costs of publishing [@Bj_rk_learned_2014]. This situation is similar to the lack of transparency regarding journal subscriptions [@Lawson_Meghreblian_2015]. It also remains unclear which factors contribute to price formation.[^4b] While fixed prices for individual articles are common, agreements between publishers and institutions often provide discounts and publishers sometimes waive publication fees for authors from low-income countries [@BJ_RK_2012; @Lawson_waiver_2015]. Other factors leading to a complex landscape of variable pricing schemes [@Pinfield_2015] include submission or page charges [@BJ_RK_2012].

Hybrid journals substantially add to the complexity of open access funding. These journals, allowing articles to be published immediately as open access after a charge was paid, rely both on subscriptions and publication fees as revenue sources, which, in turn, leads to the phenomena of "double dipping" in academic publishing [@Pinfield_2015]. Although the uptake of open access through hybrid journals were described as low [@Solomon_2012; @Bj_rk_learned_2014], this model has gained attention through recent science policies, notably because of open access policies from the UK [@Pinfield_Review_2015].

This complex situation of fee-based open access publishing creates difficulties for researchers and practitioners alike. Research addressing the question about how much was spent on open access journal articles either surveyed authors [@soap], collected price information from journal websites [@Walters_2011, @Solomon_2012], or used self-reported data [@Pinfield_2015]. In the latter case, some research funders have begun collecting and disclosing expenditures for open access journal articles as open data.

To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc] and the Austrian Science Fund FWF [@fwf_apc]. The not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report expenditures along with bibliographic information, including title, journal and publisher, persistent identifier to the publisher's version, and a link to a deposit in a subject repository. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates. In the case of the Wellcome Trust, crowd-sourcing data cleaning activities through a Google spreadsheet in combination with checks against bibliographic sources massively improved the spending data (see comments in @wellcome_apc).

In examining self-reported institutional payments made for open access articles in Germany, this paper seeks to contribute to the evolving empirical evidence on questions about open access publication fees as business model in academic publishing. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has been encouraging open access publishing since years. The DFG launched its "Open-Access Publishing" program in 2009. This programme has strongly influenced how university libraries support open access publication fees through funds.[^3] To reduce administrative burdens, grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. They include central invoicing schemes and related agreements between university libraries and publishers. Grantees are also required to report the institutional publication output and their publication fee spending to the DFG on a regular basis, and to present the university-wide strategy to sustain the funds when DFG’s initial support runs out lately in 2019.

The DFG enforces a set of criteria grantees have to comply with, leading to similar implementations for supporting open access publishing across German universities: these criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 € (excluding VAT) [@Fournier_2013]. Research institutes organized in the Fraunhofer-Gesellschaft, Helmholtz-Gemeinschaft, Leibniz-Gemeinschaft, and Max-Planck-Gesellschaft are not eligible for this funding program, contributing to the diversity of schemes in Germany. In response, some organizations have adopted similar processes to support authors. The Max-Planck-Gesellschaft operates their long-lasting open access activities, including handling spending and publisher agreements centrally, through the Max Planck Digital Library [@Schimmer_2013; @Sikora_2015], while the Leibniz-Gemeinschaft set up a dedicated open access fund in 2016.

The evolving institutional support structures to cover open access publication fees has led to calls for an unified approach towards supporting open access journal publishing. The Allianz der Wissenschaftsorganisationen[^10c], representing all major research organizations in Germany, marks in this regard transparency as one way to sustain an "adequate open access publication system" [@allianz]. Reflecting Austrian and UK initiatives to share institutional spending on open access publication fees as open data, as well as professional discussions on open access publishing, Bielefeld University Library began to openly share its payments in May 2014. After putting its approach to the working group "Electronic Publishing" of the Deutsche Initiative für Netzwerkinformation (DINI)[^11] as a basis for discussion, the library invited others to participate under the umbrella of the Open APC initiative, and contributions from other German research institutions followed soon after.

Drawing on self-reported data from this open data initiative, the aim of the study was to examine how much German universities and research organisations spent on open access publication fees until 2015. The analysis focused on the amount of institutional payments for publication fees, and compared these findings with those from related Austrian and UK initiatives. We also asked how thoroughly self-reported articles were indexed in CrossRef, a DOI minting agency for scholarly literature, and analysed how institutional spending per articles was distributed over publishers and journal titles.
